Challenges to transboundary fisheries management in North America under climate change

### 2021-03-22 Ecol/Manuscript_Draft.Rmd
William comments included

In 1982, the United Nations Law of the Sea Convention (UNCLOS) formalized the concept of exclusive economic zones (EEZs) creating the concept that we know today as shared stocks [@UnitedNations:1986tl], i.e., stocks that migrate between countries EEZs (known as transboundary stocks) or between EEZs and the high seas (also called straddling stocks) [@Song:2017iua]. Today, an estimated 347 [@Teh:2015gd] to 633 (Chapter 2) fish stocks have distributions that cross national borders, some of them jointly managed by two or more countries. These species are responsible for more than 70% of these countries total fish catches (Chapter 2). Under Article 63, UNCLOS incentives actions to cooperate on the management of shared stocks [@UnitedNations:1986tl] as management success often depends on effective cooperation between parties [@Munro:2002uf; @Sumaila:2013hv]. Since the definition of shared stocks, game theory has been one of the most common approaches used to analyze the management of these type of stocks. However, shared stocks' management can be convoluted due to the participation of several fishing “players”, different countries and sometimes jurisdictions within a country, the migration patterns of the stock, and their abundance fluctuation within space and time [@Munro:2002uf; @Engler2020]. In addition, international treaties might not be prepared to address the effects that climate change will bring to shared fish stocks [@Engler2020; @Kubrack2020]
<!-- ##### Climate Change and Transboundary Fisheries -->

@@

Quota allocation ruled by historic distributions will most likely be outdated incentivizing maladaptation [@FAO:2018uv; @Gaines:2018sg; @Miller:2013iv]. In Europe, for example, the EU Common Fisheries Policy quota allocation is based on historic reference period of the 70's [@Harte:2019us]. However, climate change has shifted the distribution of multiple European commercial stocks [@Baudron:2020dj], outdating the fixed quotas and thus compromising the sustainability of European fisheries [@FAO:2018uv; @Baudron:2020dj]. Management regimes that include a dynamic harvest control (e.g., adjusting the quota based on the stocks distribution) have the potential of increasing fish biomass, harvest and profits under climate change [@Gaines:2018sg]. In North America, poleward shifts of Pacific halibut along the coast of Oregon, Washington and British Columbia have been previously addressed by the IPHC resulting in the adoption of a dynamic quota allocation method [@McCaughran:1992uw]. By allocating quotas based on yearly surveys along the Convention area, the IPHC should be able to capture shifts in Pacific halibut distribution due to climate change, reducing the chances of over exploitation of the stock due to these shifts [@Miller:2013iv]. Similarly, for the Gulf of Maine, since the GoMa’s method to estimate quota allocation is weighted based on stocks distribution (90%) and historical catch (10%) [@TRAC:2016wu]. This process is especially important for cod and haddock due to their distribution variation within the Gulf [@Soboil:2006gw; @TRAC:2016wu]. However, since 2010, when the weighted method was implemented, the quota allocation has favored the US over Canada, especially in terms of haddock and yellowtail flounder [@Trinko:2019ou]. A perpetuation of this trend with no mitigation policy could jeopardize the arrangement as Canada's quota reduction could disincentive cooperation [@Sumaila:2020tp]. **Previous international action suggests that quota allocation should not only be centered in highly valuable fisheries (i.e., in terms of catch or economic value). For example, Atlantic mackerel (*Scomber scombrus*) represents a small proportion of the total catch of the EU, Norway, and the Faroe Islands (Denmark) who share the stock. Yet, they have all been in dispute with Iceland over quota allocation and rights since the stock shifted in 2017 reaching Icelandic waters [@Spijkers:2017ij, Chapter 2]. Similar, Peru has recently signed an agreement with Chile to manage the southern stock of Anchoveta (*Engraulis ringens*) which is substantially smaller than the northern Peruvian stock [@Cashion:2018cg; @UNDP:2016wp]. These examples suggest that a future change in shared proportion of even a "small" fishery, or fraction of, could still require management adaptations in order to reduce international conflict and sustain the stock. Naturally, no strategy to mitigate climate change impacts will be beneficial if a fishery is exploited to extinction, however, reformulation of fisheries management could actually offset many of the negative effects of climate change [@Gaines:2018sg]. Thus, the impacts and solutions presented here need to be align with sustainable fisheries management.**

>>>>

In 1982, the United Nations Law of the Sea Convention (UNCLOS) formalized the concept of exclusive economic zones (EEZs) creating the concept that we know today as shared stocks [@UnitedNations:1986tl], i.e., stocks that migrate between countries EEZs (known as transboundary stocks) or between EEZs and the high seas (also called straddling stocks) [@Song:2017iua]. Today, an estimated 347 [@Teh:2015gd] to 633 (Chapter 2) fish stocks have distributions that cross national borders, some of them jointly managed by two or more countries. These species are responsible for more than 70% of these countries total fish catches (Chapter 2). Under Article 63, UNCLOS incentives actions to cooperate on the management of shared stocks [@UnitedNations:1986tl] as management success often depends on effective cooperation between parties [@Munro:2002uf; @Sumaila:2013hv]. Since the definition of shared stocks, game theory has been one of the most common approaches used to analyze the management of these type of stocks. However, shared stocks' management can be convoluted due to the participation of several fishing “players”, different countries and sometimes jurisdictions within a country, the migration patterns of the stock, and their abundance fluctuation within space and time [@Munro:2002uf; @Engler:2020bg]. In addition, international treaties might not be prepared to address the effects that climate change will bring to shared fish stocks [@Engler:2020bg; @Koubrak:2020hw]
<!-- ##### Climate Change and Transboundary Fisheries -->

@@

Quota allocation ruled by historic distributions will most likely be outdated incentivizing maladaptation [@FAO:2018uv; @Gaines:2018sg; @Miller:2013iv]. In Europe, for example, the EU Common Fisheries Policy quota allocation is based on historic reference period of the 70's [@Harte:2019us]. However, climate change has shifted the distribution of multiple European commercial stocks [@Baudron:2020dj], outdating the fixed quotas and thus compromising the sustainability of European fisheries [@FAO:2018uv; @Baudron:2020dj]. Management regimes that include a dynamic harvest control (e.g., adjusting the quota based on the stocks distribution) have the potential of increasing fish biomass, harvest and profits under climate change [@Gaines:2018sg]. In North America, poleward shifts of Pacific halibut along the coast of Oregon, Washington and British Columbia have been previously addressed by the IPHC resulting in the adoption of a dynamic quota allocation method [@McCaughran:1992uw]. By allocating quotas based on yearly surveys along the Convention area, the IPHC should be able to capture shifts in Pacific halibut distribution due to climate change, reducing the chances of over exploitation of the stock due to these shifts [@Miller:2013iv]. Similarly, for the Gulf of Maine, since the GoMa’s method to estimate quota allocation is weighted based on stocks distribution (90%) and historical catch (10%) [@TRAC:2016wu]. This process is especially important for cod and haddock due to their distribution variation within the Gulf [@Soboil:2006gw; @TRAC:2016wu]. However, since 2010, when the weighted method was implemented, the quota allocation has favored the US over Canada, especially in terms of haddock and yellowtail flounder [@Trinko:2019ou]. A perpetuation of this trend with no mitigation policy could jeopardize the arrangement as Canada's quota reduction could disincentive cooperation [@Sumaila:2020tp]. Previous international action suggests that shifts in transboundary fish stocks distributions for fisheries with relatively low catch or economic value could lead to international disputes in quota allocation. For example, Atlantic mackerel (*Scomber scombrus*) represents a small proportion of the total catch of the EU, Norway, and the Faroe Islands (Denmark) who share the stock. Yet, they have all been in dispute with Iceland over quota allocation and rights since the stock shifted in 2017 reaching Icelandic waters [@Spijkers:2017ij, Chapter 2]. Similarly, Peru has recently signed an agreement with Chile to manage the southern stock of Anchoveta (*Engraulis ringens*) which is substantially smaller than the northern Peruvian stock [@Cashion:2018cg; @UNDP:2016wp]. These examples highlight the need for adaptation of transboundary fisheries management to climate change for all a broad range of species in order to reduce the risk of international conflict and unsustainable fisheries development. Naturally, no strategy to mitigate climate change impacts will be beneficial if a fishery is exploited to extinction, however, reformulation of fisheries management could actually offset many of the negative effects of climate change [@Gaines:2018sg]. Thus, the impacts and solutions presented here need to be align with sustainable fisheries management.
